Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-21 optim.qmd
update the squintability measure x4

    -   Use jellyfish search optimiser, or JS optimiser, not jellyfish
        optimiser (my bad for leading the way)
    -   "hyper-parameter" rather than "hyperparameter"

@@

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely *projection
pursuit index (PPI)* (@hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020). Given
high-dimensional data $X \in \mathbf{R}^{n }\times p$ and PPI
$f(\cdot)$, PP finds the orthonormal projection matrix
$A \in \mathbf{R}^{p \times d}$ by solving the following optimisation
problem:

@@

The primary goal of this study is to investigate the performance of JSO in the context of the PP guided tour. A series of simulation experiments using well-recognised data sets and PPIs are conducted to yield insights regarding the behaviour of JSO under different settings and the sensitivity of hyper-parameter in the optimisation. Second, to observe the performance of JSO with different types of PPI, a collection of metrics is introduced to capture specific properties of the index including squintability and smoothness as introduced in @laa_using_2020. To the best of the authors’ knowledge, this is the first attempt to quantitatively measure squintability and smoothness. Finally, the relationship between the JSO performance, hyper-parameters tuning and various properties of PPI is analysed to provide helpful guidance for practitioners that are using the guided tour for high dimensional visualisation.

The rest of this paper is structured as follows. @sec-background
introduces the background of the PP guided tour and reviews existing methods in the literature. @sec-theory describes the details of JSO
and introduces metrics that measure different properties of a PPI.
@sec-vis visualises the behaviour of the JSO in terms of describes a simulation study on the performance of the JSO using well-known data sets and index functions. @sec-app displays two sets of simulation experiments comparing the JSO with the search-better optimiser and studies the impact of different PPI properties on the optimisation performance. @sec-conclusion summarises the work and provides suggestions for future directions.

@@

The jellyfish hyper-parameters used are 50 jellyfishes (`n_jellies`) and
a maximum of 50 tries (`max_tries`). The full data contains 50
simulations, recorded by the variable `sim`, with different seeds
(`seed`). Recorded in columns `basis`, `index_val`, `tries`, `loop` and
`time`, are the sampled projection basis, calculated index value,
iteration ID, jellyfish ID, and time taken to find the optimum,
respectively. The basis column records every visited basis before
comparing with the current (last) basis. If a visited basis has a
smaller index value than the current (last) one, the JS optimiser keeps
the current basis for the next iteration in the search but still records
the visited inferior basis. This data can be used to derive numerical
summaries and plots:

@@

The JS optimiser provides a new swarm-based search strategy to optimise
projection pursuit problems. Two examples are provided in this section
to demonstrate its performance. The first example compares its
performance with the search-better optimiser and explores how it behaves
with different combinations of hyper-parameters. The second example
studies the effect of optimisation properties defined in @sec-theory,
along with jellyfish hyper-parameters, on the outcome of the
optimisation.

## Performance of the JS optimiser in pipe-finding problems {#sec-app-1}

The performance of the JS optimiser is investigated in two folds: 1)
compare it with a commonly used optimiser: the search-better optimiser
[@RJ-2021-105; @laa_using_2020], and 2) examine its success rate under
different hyper-parameters (number of jellyfishes and maximum number of
tries). The projection pursuit problem used is finding the pipe shape
using the holes index, investigated by @laa_using_2020.

```{r}
#| fig.height = 7.5,
#| fig.width = 10,
#| label = fig-proj,
#| fig.align = "center",
#| fig.cap = "Projections found by the jellyfish and search better optimisers at each 10th quantile across 50 simulations. The projection pursuit problem is to find the pipe shape using the holes index in the 6, 8, 10, and 12-dimensional spaces. The JS optimiser uses 100 jellyfishes and a maximum number of tries of 100. The search better optimiser uses a maximum of 1000 tries in each step of random sampling step before the algorithm terminates. In the 6-D data space, the JS optimiser always finds a clear pipe shape while the search better optimiser also finds the pipe shape but with a wide rim. At higher data dimensions, the JS optimiser finds a higher index value and a clearer pipe shape across all the quantiles than the search better optimiser."

@@

(p1 + ggtitle("The JS optimiser"))/(p2 + ggtitle("The search better optimiser"))
```

@@

Fifty simulations are conducted with both JS and the search-better
optimiser, in four dimensions ($d = 6, 8, 10, 12$). The JS optimiser
uses 100 jellyfishes and 100 maximum number of tries, while the
search-better optimiser allows a maximum of 1000 samples at each
iteration before the algorithm terminates. @fig-proj presents the final
projections found by the two optimisers, broken down by 10th quantile,
faceted by the data dimension. In the 6-dimensional data scenario, the
JS optimiser consistently identifies a clear pipe shape. The
search-better optimiser also finds the pipe shape but with a wide rim,
suggesting a further polish search may be required. With increasing
dimensions, the JS optimiser may not always identify the pipe shape due
to random sampling, but it still presents a pipe shape in over 50% of
cases. When compared to the search-better optimiser, the JS optimiser
reaches higher index values and clearer pipe shapes across all
quantiles.

In the second experiment, fifty simulations are conducted at each
hyper-parameter combination to analyse its effects on the success rate
of the JS optimiser. The hyper-parameters tested include: 1) 20, 50, and
100 jellyfishes, and 2) 50 and 100 maximum number of tries. In each
simulation, the success rate is calculated as defined in @sec-vis.
@fig-proportion presents the proportion of success for each
hyper-parameter combinations. As the number of jellyfishes and maximum
tries increase, the success rate also increases. For simpler problems (6
dimensions), small parameter values (20 jellyfishes and a maximum number
of tries of 50) can already result in high success rate, while for
higher-dimensional problems (8, 10, and 12 dimensions), a combination of
100 jellyfishes and 100 maximum number of tries is necessary for at
least half of the jellyfishes to find the optimal projection.

## Factors affecting the JS optimiser success rate: optimisation properties and jellyfish hyper-parameters {#sec-app-2}

The optimisation properties, including smoothness, squintability, and
speed, offer numerical metrics to characterise the complexity of a
projection pursuit optimisation problem. This example investigates how
these metrics, along with the jellyfish search hyper-parameters (the
number of jellyfishes and the maximum number of tries), affect the
success of the JS optimiser. Simulations are conducted to obtain the
success rate across various projection pursuit problems, characterised
by shape-to-find, data dimension, and index function used, as well as
different hyper-parameter combinations. Smoothness and squintability are
calculated for each projection pursuit problem as outlined in
@sec-theory. A generalised linear model is used to construct the
relationship between the success rate, jellyfish hyper-parameters and
optimisation properties.

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indices considered:
`dcor2d_2`, `loess2d`, `MIC`, `TIC`, `spline`, and `stringy`. Combining
with two jellyfish hyper-parameters, a total of `r nrow(sim_df)` cases
is produced, comprising of `r nrow(filter(sim_df, index == "holes"))`
pipe-finding cases and `r nrow(filter(sim_df, index != "holes"))`
sine-wave finding cases. For each case, the JS optimiser is run fifty
times and the success rate is calculated as in @sec-app-1.

@@

@tbl-mod-data combines all the variables calculated above (the JS
optimiser success rate, jellyfish hyper-parameters, and optimisation
properties) into one table. Three preprocessing steps are applied to the
data: 1) scaling the jellyfish hyper-parameters by a factor of 10 for
interpretation, 2) creating a new binary variable `long_time` to
indicate cases with an average run time over 30 seconds, and 3) encoding
the success rate for the stringy case as 0, since none of the 50
simulations identified the sine-wave shape. A generalised linear model
with a binomial family and a logit link function is used to fit the data
and @tbl-mod-output presents the model outputs. The model suggests that
the success rate of the JS optimiser is positively associated with the
two jellyfish hyper-parameters, as well as with smoothness and
squintability. However, being flagged with long runtime and an increase
of data dimension will reduce the success rate. The variable
`squintability` and `dimension` are significant, suggesting their
importance relative to jellyfish hyper-parameters in the optimisation
success.

@@

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with the JS optimiser success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
sim_df2 <- sim_df |>   
  mutate(n_jellies = n_jellies/10,
         max_tries = max_tries/10,
         long_time = ifelse(time > 20, 1, 0),
         P_J = ifelse(index == "stringy", 0, P_J))

>>>>

    -   Use jellyfish search optimiser, or JSO
    -   "hyper-parameter" rather than "hyperparameter"
    - PP index rather than PPI

@@

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely *PP index* 
(@hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020). Given
high-dimensional data $X \in \mathbf{R}^{n }\times p$ and the index function
$f(\cdot)$, PP finds the orthonormal projection matrix
$A \in \mathbf{R}^{p \times d}$ by solving the following optimisation
problem:

@@

The primary goal of this study is to investigate the performance of JSO in the context of the Projection pursuit. A series of simulation experiments using well-recognised data sets and PP indexes are conducted to yield insights regarding the behaviour of JSO under different settings and the sensitivity of hyper-parameter in the optimisation. Second, to observe the performance of JSO with different types of PP indexes, a collection of metrics is introduced to capture specific properties of the index including squintability and smoothness as introduced in @laa_using_2020. To the best of the authors’ knowledge, this is the first attempt to quantitatively measure squintability and smoothness. Finally, the relationship between the JSO performance, hyper-parameters tuning and various properties of PP indexes is analysed to provide helpful guidance for practitioners that are using the guided tour for high dimensional visualisation.

The rest of this paper is structured as follows. @sec-background
introduces the background of the PP guided tour and reviews existing methods in the literature. @sec-theory describes the details of JSO
and introduces metrics that measure different properties of PP indexes.
@sec-vis visualises the behaviour of the JSO in terms of describes a simulation study on the performance of the JSO using well-known data sets and index functions. @sec-app displays two sets of simulation experiments comparing the JSO with the search-better optimiser and studies the impact of different PP index properties on the optimisation performance. @sec-conclusion summarises the work and provides suggestions for future directions.

@@

The jellyfish hyper-parameters used are 50 jellyfishes (`n_jellies`) and
a maximum of 50 tries (`max_tries`). The full data contains 50
simulations, recorded by the variable `sim`, with different seeds
(`seed`). Recorded in columns `basis`, `index_val`, `tries`, `loop` and
`time`, are the sampled projection basis, calculated index value,
iteration ID, jellyfish ID, and time taken to find the optimum,
respectively. The basis column records every visited basis before
comparing with the current (last) basis. If a visited basis has a
smaller index value than the current (last) one, JSO will keep
the current basis for the next iteration in the search but still records
the visited inferior basis. This data can be used to derive numerical
summaries and plots:

@@

The JSO provides a new swarm-based search strategy to optimise
projection pursuit problems. Two examples are provided in this section
to demonstrate its performance. The first example compares its
performance with the search-better optimiser and explores how it behaves
with different combinations of hyper-parameters. The second example
studies the effect of optimisation properties defined in @sec-theory,
along with jellyfish hyper-parameters, on the outcome of the
optimisation.

## Performance of JSO in pipe-finding problems {#sec-app-1}

The performance of JSO is investigated in two folds: 1)
compare it with a commonly used optimiser: the search-better optimiser
[@RJ-2021-105; @laa_using_2020], and 2) examine its success rate under
different hyper-parameters (number of jellyfishes and maximum number of
tries). The projection pursuit problem used is finding the pipe shape
using the holes index, investigated by @laa_using_2020.

```{r}
#| fig.height = 7.5,
#| fig.width = 10,
#| label = fig-proj,
#| fig.align = "center",
#| fig.cap = "Projections found by the jellyfish and search better optimisers at each 10th quantile across 50 simulations. The projection pursuit problem is to find the pipe shape using the holes index in the 6, 8, 10, and 12-dimensional spaces. The JSO uses 100 jellyfishes and a maximum number of tries of 100. The search better optimiser uses a maximum of 1000 tries in each step of random sampling step before the algorithm terminates. In the 6-D data space, JSO always finds a clear pipe shape while the search better optimiser also finds the pipe shape but with a wide rim. At higher data dimensions, JSO finds a higher index value and a clearer pipe shape across all the quantiles than the search better optimiser."

@@

(p1 + ggtitle("The jellyfish search optimiser"))/(p2 + ggtitle("The search better optimiser"))
```

@@

Fifty simulations are conducted with both JSO and the search-better
, in four dimensions ($d = 6, 8, 10, 12$). The JSO
uses 100 jellyfishes and 100 maximum number of tries, while the
search-better optimiser allows a maximum of 1000 samples at each
iteration before the algorithm terminates. @fig-proj presents the final
projections found by the two optimisers, broken down by 10th quantile,
faceted by the data dimension. In the 6-dimensional data scenario, JSO
 consistently identifies a clear pipe shape. The
search-better optimiser also finds the pipe shape but with a wide rim,
suggesting a further polish search may be required. With increasing
dimensions, JSO may not always identify the pipe shape due
to random sampling, but it still presents a pipe shape in over 50% of
cases. When compared to the search-better optimiser, JSO
reaches higher index values and clearer pipe shapes across all
quantiles.

In the second experiment, fifty simulations are conducted at each
hyper-parameter combination to analyse its effects on the success rate
of JSO . The hyper-parameters tested include: 1) 20, 50, and
100 jellyfishes, and 2) 50 and 100 maximum number of tries. In each
simulation, the success rate is calculated as defined in @sec-vis.
@fig-proportion presents the proportion of success for each
hyper-parameter combinations. As the number of jellyfishes and maximum
tries increase, the success rate also increases. For simpler problems (6
dimensions), small parameter values (20 jellyfishes and a maximum number
of tries of 50) can already result in high success rate, while for
higher-dimensional problems (8, 10, and 12 dimensions), a combination of
100 jellyfishes and 100 maximum number of tries is necessary for at
least half of the jellyfishes to find the optimal projection.

## Factors affecting JSO success rate: optimisation properties and jellyfish hyper-parameters {#sec-app-2}

The optimisation properties, including smoothness, squintability, and
speed, offer numerical metrics to characterise the complexity of a
projection pursuit optimisation problem. This example investigates how
these metrics, along with the jellyfish search hyper-parameters (the
number of jellyfishes and the maximum number of tries), affect the
success of JSO. Simulations are conducted to obtain the
success rate across various projection pursuit problems, characterised
by shape-to-find, data dimension, and index function used, as well as
different hyper-parameter combinations. Smoothness and squintability are
calculated for each projection pursuit problem as outlined in
@sec-theory. A generalised linear model is used to construct the
relationship between the success rate, jellyfish hyper-parameters and
optimisation properties.

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indices considered:
`dcor2d_2`, `loess2d`, `MIC`, `TIC`, `spline`, and `stringy`. Combining
with two jellyfish hyper-parameters, a total of `r nrow(sim_df)` cases
is produced, comprising of `r nrow(filter(sim_df, index == "holes"))`
pipe-finding cases and `r nrow(filter(sim_df, index != "holes"))`
sine-wave finding cases. For each case, JSO is run fifty
times and the success rate is calculated as in @sec-app-1.

@@

@tbl-mod-data combines all the variables calculated above (the JS
optimiser success rate, jellyfish hyper-parameters, and optimisation
properties) into one table. Three preprocessing steps are applied to the
data: 1) scaling the jellyfish hyper-parameters by a factor of 10 for
interpretation, 2) creating a new binary variable `long_time` to
indicate cases with an average run time over 30 seconds, and 3) encoding
the success rate for the stringy case as 0, since none of the 50
simulations identified the sine-wave shape. A generalised linear model
with a binomial family and a logit link function is used to fit the data
and @tbl-mod-output presents the model outputs. The model suggests that
the success rate of JSO is positively associated with the
two jellyfish hyper-parameters, as well as with smoothness and
squintability. However, being flagged with long runtime and an increase
of data dimension will reduce the success rate. The variable
`squintability` and `dimension` are significant, suggesting their
importance relative to jellyfish hyper-parameters in the optimisation
success.

@@

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with JSO success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
sim_df2 <- sim_df |>   
  mutate(n_jellies = n_jellies/10,
         max_tries = max_tries/10,
         long_time = ifelse(time > 20, 1, 0),
         P_J = ifelse(index == "stringy", 0, P_J))
